\documentclass[11pt,a4paper]{book}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[top=2.5cm, bottom=1.5cm, left=1.5cm, right=1.5cm]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\allowdisplaybreaks
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage[most]{tcolorbox}
\usepackage{fancyhdr}
\usepackage{array}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{float}
\usepackage{calligra}
\usepackage[export]{adjustbox}
\usepackage[font=small, labelfont=bf]{caption}
\usepackage{titlesec}

\usetikzlibrary{calc}

\definecolor{primaryblue}{HTML}{6BA3C8}
\definecolor{secondaryblue}{HTML}{A8D8E8}
\definecolor{lightgray}{HTML}{F5F5F0}
\definecolor{darkgray}{HTML}{3D3D3D}
\definecolor{dangerred}{HTML}{E8B4B8}
\definecolor{navyblue}{HTML}{4A3C3A}
\definecolor{chaptercolor}{HTML}{2F6FA3}
\definecolor{sectioncolor}{HTML}{5C524F}
\definecolor{subsectioncolor}{HTML}{C46A75}
\definecolor{subsubsectioncolor}{HTML}{7C68A6}
\definecolor{softlavender}{HTML}{D4C5E8}
\definecolor{softpeach}{HTML}{F4C2A3}
\definecolor{mintgreen}{HTML}{A8D5BA}
\definecolor{softcoral}{HTML}{F5A5A8}
\setlength{\headheight}{24pt}
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{6pt}
\newcolumntype{C}[1]{>{\centering\arraybackslash}m{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}m{#1}}
\setlength{\LTleft}{\fill}
\setlength{\LTright}{\fill}

\renewcommand{\chaptermark}[1]{\markboth{Chapter \thechapter: #1}{}}
\renewcommand{\sectionmark}[1]{\markright{Section \thesection: #1}}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{%
    \begin{minipage}[t]{0.48\textwidth}
    \raggedright
    \adjustbox{max width=\linewidth}{\textcolor{darkgray}{\textbf{\leftmark}}}%
    \end{minipage}}
\fancyhead[R]{%
    \begin{minipage}[t]{0.48\textwidth}
    \raggedleft
    \adjustbox{max width=\linewidth}{\textcolor{darkgray}{\textbf{\rightmark}}}%
    \end{minipage}}
\fancyfoot[C]{\textcolor{darkgray}{\thepage}}
\fancyfoot[R]{\tiny\textcalligra{\textcolor{dangerred}{Miad}}}
\renewcommand{\headrulewidth}{2pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{primaryblue}\leaders\hrule height \headrulewidth\hfill}}

\fancypagestyle{plain}{%
    \fancyhf{}
    \fancyfoot[C]{\textcolor{darkgray}{\thepage}}
    \fancyfoot[R]{\tiny\textcalligra{\textcolor{dangerred}{Miad}}}
    \renewcommand{\headrulewidth}{0pt}
}

\setcounter{secnumdepth}{5}

\titleformat{\chapter}
  {\normalfont\Huge\bfseries\color{chaptercolor}}
  {Chapter \thechapter.}
  {1em}
  {}
  [\vspace{2pt}\textcolor{chaptercolor}{\titlerule[2.5pt]}]

\titleformat{\section}
  {\normalfont\huge\bfseries\color{sectioncolor}}
  {\thesection}
  {1em}
  {}

\titleformat{\subsection}
  {\normalfont\LARGE\bfseries\color{subsectioncolor}}
  {\thesubsection}
  {1em}
  {}

\titleformat{\subsubsection}
  {\normalfont\Large\bfseries\color{subsubsectioncolor}}
  {}
  {0pt}
  {\thesubsubsection\quad}

\titleformat{name=\subsubsection,numberless}
  {\normalfont\Large\bfseries\color{subsubsectioncolor}}
  {}
  {0pt}
  {}

\titleformat{\paragraph}[block]
  {\normalfont\large\bfseries\color{sectioncolor}}
  {}
  {0pt}
  {\theparagraph\quad}

\titleformat{name=\paragraph,numberless}[block]
  {\normalfont\large\bfseries\color{sectioncolor}}
  {}
  {0pt}
  {}

\titleformat{\subparagraph}[block]
  {\normalfont\normalsize\bfseries\color{darkgray}}
  {}
  {0pt}
  {\thesubparagraph\quad}

\titleformat{name=\subparagraph,numberless}[block]
  {\normalfont\normalsize\bfseries\color{darkgray}}
  {}
  {0pt}
  {}

\titlespacing*{\chapter}{0pt}{0pt}{2.3ex plus .2ex}
\titlespacing*{\section}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}
\titlespacing*{\subsection}{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\titlespacing*{\subsubsection}{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\titlespacing*{\paragraph}{0pt}{2.5ex plus .5ex minus .2ex}{0.8ex}
\titlespacing*{\subparagraph}{0pt}{1.5ex plus .3ex minus .2ex}{0.5ex}

\newenvironment{definitionbox}[1][Definition]{%
    \begin{tcolorbox}[
        enhanced,
        breakable,
        colback=primaryblue!10,
        colframe=primaryblue!150,
        title=#1,
        fonttitle=\bfseries,
        sharp corners,
        boxrule=0.8pt,
        left=8pt,
        right=8pt,
        top=6pt,
        bottom=6pt
    ]
}{\end{tcolorbox}}

\newenvironment{formulabox}[1][Formula]{%
    \begin{tcolorbox}[
        enhanced,
        breakable,
        colback=softcoral!10,
        colframe=softcoral!200,
        title=#1,
        fonttitle=\bfseries,
        coltitle=white,
        sharp corners,
        boxrule=1pt,
        left=8pt,
        right=8pt,
        top=6pt,
        bottom=6pt
    ]
}{\end{tcolorbox}}

\newenvironment{infobox}[1][Information]{%
    \begin{tcolorbox}[
        enhanced,
        breakable,
        colback=mintgreen!10,
        colframe=mintgreen!200,
        title=#1,
        fonttitle=\bfseries,
        sharp corners,
        boxrule=0.8pt,
        left=8pt,
        right=8pt,
        top=6pt,
        bottom=6pt
    ]
}{\end{tcolorbox}}

\newcounter{qna}[chapter]
\renewcommand{\theqna}{\thechapter.\arabic{qna}}
\newenvironment{questionanswerbox}[1][Example]{%
    \refstepcounter{qna}
    \begin{tcolorbox}[
        enhanced,
        breakable,
        colback=softlavender!10,
        colframe=softlavender!200,
        title={#1~\theqna},
        fonttitle=\bfseries,
        coltitle=white,
        colbacktitle=softlavender!200,
        sharp corners,
        boxrule=1pt,
        left=8pt,
        right=8pt,
        top=6pt,
        bottom=6pt
    ]
}{\end{tcolorbox}}

\setlist[itemize,1]{label=\textcolor{primaryblue}{$\triangleright$}, leftmargin=*, itemsep=4pt, topsep=4pt}
\setlist[enumerate,1]{label=\textcolor{primaryblue}{\arabic*.}, leftmargin=*, itemsep=5pt, topsep=4pt}

\hypersetup{
    colorlinks=true,
    linkcolor=primaryblue,
    urlcolor=primaryblue,
    citecolor=primaryblue
}

\begin{document}
\setcounter{page}{1}
\tableofcontents
\clearpage

\chapter{Introduction}

% Source PDF page 5
\section{Information}

\begin{definitionbox}[Information]
Information refers to data or knowledge that is meaningful and has value in making decisions or solving problems.
\end{definitionbox}

\subsection{Importance of Information Theory}

\subsubsection{Why Information Theory Matters}
\begin{itemize}
    \item It helps us \textbf{quantify} information.
    \item It helps us \textbf{transmit} information efficiently.
    \item It helps us \textbf{optimize the delivery} of information in communication, computing, and biological systems.
\end{itemize}

\subsubsection{Real-Life Relevance}
\begin{itemize}
    \item Modern systems such as the \textbf{internet}, \textbf{mobile networks}, and \textbf{social media} rely on efficient data compression and error correction.
    \item These ideas ensure messages are transmitted \textbf{accurately} and \textbf{efficiently}.
    \item Information theory also helps reduce \textbf{noise interference} in signals.
    \item This improves reliability in applications such as \textbf{streaming}, \textbf{medical diagnostics}, and \textbf{genomic analysis}.
\end{itemize}

\subsection{Importance in Communication}

\subsubsection{Key Points}
\begin{itemize}
    \item Information theory provides mathematical frameworks to optimize \textbf{data transmission} and \textbf{storage}.
    \item It helps determine the maximum amount of information a channel can handle, that is, \textbf{channel capacity}.
    \item It helps determine how to encode messages so that transmission errors can be reduced.
\end{itemize}

\subsubsection{Communication Insight}
\begin{itemize}
    \item Cellular networks and internet protocols use \textbf{error correction} and \textbf{data compression}.
    \item These make communication \textbf{efficient} and \textbf{reliable}.
    \item Even in noisy environments, users can still experience:
    \begin{itemize}
        \item clear calls,
        \item fast internet speeds,
        \item accurate data exchange.
    \end{itemize}
\end{itemize}

\noindent This is why information theory is fundamental to seamless global communication.

\clearpage

% Source PDF page 6
\section{Discrete Source}

\begin{definitionbox}[Discrete Source]
A discrete source is a system that generates a \textbf{finite set of symbols or values} from a \textbf{predefined set of alphabets}. These symbols are discrete and distinct, such as letters, numbers, or pixels.
\end{definitionbox}

\subsubsection{Examples of Discrete Sources}
\begin{itemize}
    \item \textbf{Text:} A keyboard generates letters, numbers and special characters.
    \item \textbf{Audio:} Digital audio files sample sound at specific intervals, producing discrete data points.
    \item \textbf{Image/Video:} Images are composed of pixels with distinct color values, while videos are sequences of discrete image frames.
\end{itemize}

\subsection{Source Types}
\subsubsection{Memoryless Source}

\begin{definitionbox}[Memoryless Source]
A discrete source is memoryless if the symbols it generates are \textbf{independent} of each other.
\end{definitionbox}

\begin{itemize}
    \item The current symbol has \textbf{no dependence} on past symbols.
    \item Example: flipping a fair coin repeatedly produces independent outcomes $(H \text{ or } T)$.
    \item Such sources simplify analysis because each symbol can be treated individually without considering historical context.
    \item Example:A random number generator producing digits $(0\text{--}9)$ where each number is equally likely and independent of the previous numbers.
\end{itemize}

\subsubsection{Memory Source}

\begin{definitionbox}[Memory Source]
A discrete source is a memory source if the generation of a symbol \textbf{depends} on one or more \textbf{previous symbols}.
\end{definitionbox}

\begin{itemize}
    \item Such sources have \textbf{statistical dependency} between symbols.
    \item This dependency can be modeled using \textbf{Markov processes} or \textbf{conditional probabilities}.
    \item They are common in \textbf{natural language and encoded data streams}, where rules and structure influence symbol generation.
    \item Example: In English text, the letter `q' is almost always followed by `u', showing a strong dependency between these symbols.
\end{itemize}
\clearpage

% Source PDF page 7
\section{Self Information}

\begin{definitionbox}[Self Information]
Self information measures the \textbf{amount of uncertainty or surprise} associated with the occurrence of a specific event. It quantifies how much \textbf{information is gained} when an event occurs.
\end{definitionbox}

\begin{formulabox}[Self Information Formula]
Consider a discrete random variable $X$ with possible outcomes $x_i$, $i = 1, 2, 3, \dots, n$. Self information of the event $X = x_i$ is defined as
\[
I(x_i) = \log \frac{1}{P(x_i)} = -\log P(x_i).
\]
\begin{itemize}
    \item $I(x_i)$ = self information of the event $x_i$
    \item $P(x_i)$ = probability of the event $x_i$
\end{itemize}
\end{formulabox}

\subsection{Properties of Self Information}
\begin{enumerate}
    \item If we are absolutely certain of the outcome of an event even before it occurs, there is no information gained.
    \[
    I(x_i) = 0, \quad \text{if } P_i = 1
    \]
    This means \textbf{certainty carries no message information}.
    \item The occurrence of an event either provides some or no information, but never brings about a loss of information.
    \[
    I(x_i) \geq 0, \quad \text{for } 0 \leq P(x_i) \leq 1
    \]
    That implies \textbf{self information cannot be negative}.
    \item The \textbf{less probable an event is, the more information we gain} when it occurs.
    \[
    I(x_1) > I(x_2) \quad \text{if } P_1 < P_2
    \]
    \item For statistically independent events, the total information obtained from the joint occurrence is equal to the sum of the individual informations.
    \[
    I(x_1, x_2) = I(x_1) + I(x_2).
    \]
    This means that \textbf{information is additive for independent events}.

    Equivalently,
    \[
    I_A + I_B = \log \frac{1}{P_A} + \log \frac{1}{P_B}
    = \log \frac{1}{P_A P_B}
    = \log \frac{1}{P(AB)}.
    \]

    This follows because for independent events,
    \[
    P(AB) = P_A \cdot P_B.
    \]

    Thus, the information of the joint event can be written as
    \[
    I(AB) = \log \frac{1}{P(AB)},
    \]
    which matches the sum of the individual informations. This implies that \textbf{independent events do not share information, so their informations simply add up}.
\end{enumerate}

\subsection{Physical significance / insight of self information}
Consider a set of events $X = \{x_1, x_2, x_3\}$ with probability
\[
P = \{0.5, 0.25, 0.25\}.
\]

\textbf{Step 1:} Self information for the first event:
\[
I_{x_1} = \log_2 \frac{1}{0.5} = -\log_2 0.5 = 1 \text{ bit}
\]

\textbf{Step 2:} For the other two events:
\[
I_{x_2} = I_{x_3} = \log_2 \frac{1}{0.25} = -\log_2 0.25 = 2 \text{ bit}
\]

\begin{infobox}[Observation]
\begin{itemize}
    \item Higher probability requires fewer bits.
    \item Lower probability requires more bits.
    \item Lower probability therefore means a higher degree of uncertainty.
\end{itemize}
\end{infobox}

\clearpage

% Source PDF page 8
\section{Entropy}

\begin{definitionbox}[Entropy]
In information theory, entropy is the measurement of the average amount of information, that is, the number of bits of uncertainty, in a set of possible outcomes per symbol.
\end{definitionbox}

\subsection{Interpretation}
\begin{itemize}
    \item It measures the \textbf{unpredictability} of a random variable.
    \item It represents the minimum number of bits required to encode the information generated by a source.
\end{itemize}

\begin{formulabox}[Entropy of an Event]
\[
H(x_i) = -P_i \log_2 P_i.
\]
\begin{itemize}
    \item $H(x_i)$ = entropy contribution of the event $x_i$
    \item $P_i$ = probability of the $i$th event
\end{itemize}
\end{formulabox}

\subsection{Why This Form Appears}
\begin{itemize}
    \item Self information of an event is $\left(-\log_2 P_i\right)$.
    \item Entropy is an \textbf{average} information measure.
    \item So self information is weighted by its probability $(P_i)$.
\end{itemize}

\begin{formulabox}[Entropy of a Random Variable]
\[
H(X) = -\sum_{i=1}^{n} P_i \log_2 P_i
\quad \text{bit/symbol.}
\]
\begin{itemize}
    \item $H(X)$ = average entropy of the random variable $X$
    \item $n$ = total number of possible outcomes
    \item $P_i$ = probability of the $i$th outcome
\end{itemize}
\end{formulabox}

\begin{questionanswerbox}[Example]
A source generates information with probabilities
\[
P_1 = 0.1,\quad P_2 = 0.2,\quad P_3 = 0.3,\quad P_4 = 0.4.
\]
Find the entropy of the system and what percentage of maximum possible information is being generated by this source.

\textbf{Step 1: Calculate entropy}
\[
H(X) = -\sum_{i=1}^{n} P_i \log_2 P_i
\]
\[
= -\Bigl(0.1\log_2(0.1) + 0.2\log_2(0.2) + 0.3\log_2(0.3) + 0.4\log_2(0.4)\Bigr)
\]
\[
= 1.8464 \approx 2 \text{ bits/symbol.}
\]

\textbf{Step 2: Calculate maximum entropy}
\[
H_{\max} = \log_2 4 = 2.
\]

\textbf{Step 3: Calculate efficiency}
\[
\eta = \frac{H}{H_{\max}} \times 100
= \frac{1.8464}{2}\times 100
= 92.32\%.
\]
\end{questionanswerbox}

\begin{questionanswerbox}[Question]
\textbf{Q:} Why is \texorpdfstring{$P_i$}{Pi} multiplied with self information in the entropy equation?

\textbf{A:}
\begin{itemize}
    \item Entropy is an \emph{average} measure of information.
    \item So the self information term $\left(-\log_2 P_i\right)$ must be weighted by its probability $P_i$.
    \item This keeps highly probable events contributing more frequently and less probable events contributing less frequently to the final average.
    \item It also keeps the entropy value bounded properly for a valid probability distribution.
            For example- the probability must be in between 0 and 1. If - 
            \begin{align*}
                P = 0.1 &\Rightarrow I = -\log_2(0.1) = 3.32 \text{ bits} \text{ (more than 1 bit)} \\
                P = 0.5 &\Rightarrow I = -\log_2(0.6) = 0.74 \text{ bit} \\
                P = 0.9 &\Rightarrow I = -\log_2(0.9) = 0.152 \text{ bits}
            \end{align*}
            To maintain entropy value below 1 bit/symbol, we need to multiply the self information with the probability $P_i$.
\end{itemize}
\end{questionanswerbox}

\clearpage

% Source PDF page 9
\subsection{Significance of Entropy}

\begin{itemize}
    \item \textbf{Measures uncertainty:} quantifies unpredictability in a source, providing insight into the information content.
    \item \textbf{Data compression:} sets the theoretical limits for lossless compression, helping design efficient encoding systems.
    \item \textbf{Error detection/correction:} guides the design of coding techniques to handle noise in communication channels.
    \item \textbf{System efficiency:} helps evaluate and optimize the performance of communication and storage systems.
    \item \textbf{Applications in AI/ML:} used in decision trees and deep learning for feature selection and model optimization.
\end{itemize}

\subsection{Relation between Hartley, Nats and Bits}

\subsubsection{Units of Self Information}
\begin{itemize}
    \item Self information in Hartley:
    \[
    I = \log_{10}\frac{1}{P}
    \]
    \item Self information in nats:
    \[
    I = \log_e \frac{1}{P}
    \]
    \item Self information in bits:
    \[
    I = \log_2 \frac{1}{P}
    \]
\end{itemize}

\begin{formulabox}[Hartley to Bits]
\[
1 \text{ Hartley}
= \frac{I}{-\log_{10} P}
= \frac{-\log_2 P}{-\log_{10} P}
= \frac{\log 10}{\log 2}
= \log_2 10
= 3.3219 \text{ bits.}
\]
\end{formulabox}

\begin{formulabox}[Hartley to Nats]
\[
1 \text{ Hartley}
= \frac{I}{-\log_{10} P}
= \frac{-\log_e P}{-\log_{10} P}
= \frac{\log 10}{\log e}
= \log_e 10
= 2.302 \text{ nats.}
\]
\end{formulabox}

\begin{formulabox}[Nats to Bits]
\[
1 \text{ Nat}
= \frac{I}{-\log_e P}
= \frac{-\log_2 P}{-\log_e P}
= \frac{\log e}{\log 2}
= \log_2 e
= \frac{1}{\ln 2}
= 1.442 \text{ bits.}
\]
\end{formulabox}

\begin{questionanswerbox}[Example]
\textbf{Q:} If there are $100$ equiprobable outcomes, calculate the corresponding values in nats, bits and Hartley.

\textbf{A:} Given $N = 100$, for equiprobable outcomes:
\[
P(x_i) = \frac{1}{100}
\]

\textbf{Using entropy formula:}
\[
H(X) = -\sum_{i=1}^{100} \frac{1}{100} \log \frac{1}{100}
\]

Since all terms are identical:
\[
H(X) = -100 \cdot \frac{1}{100} \log \frac{1}{100}
\]

\[
H(X) = -\log \frac{1}{100}
\]

Now simplify:
\[
H(X) = \log 100
\]

\textbf{Concept:}
\begin{itemize}
    \item Bits, nats, and Hartley are \textbf{units of information} (log base), used for both $I(x)$ and $H(X)$.
    \item The question refers to a \textbf{set of outcomes}, so we use \textbf{entropy}:
    \[
    H(X) = -\sum P(x_i)\log P(x_i)
    \]
    \item For equiprobable case:
    \[
    I(x_i) = -\log \frac{1}{100} = \log 100, \quad H(X) = \log 100
    \]
    \item Hence, \textbf{self-information = entropy} in this case.
    \item Units depend on log base:
    \[
    \log_2 \rightarrow \text{bits}, \quad
    \log_e \rightarrow \text{nats}, \quad
    \log_{10} \rightarrow \text{Hartley}
    \]
\end{itemize}

\textbf{Note:} Bits, nats and Hartley are units of information (log base).

\begin{align*}
\text{In bits, } H(X) &= \log_2(100) = 6.64 \\
\text{In nats, } H(X) &= \log_e(100) = 4.605 \\
\text{In Hartley, } H(X) &= \log_{10}(100) = 2
\end{align*}
\end{questionanswerbox}

\clearpage

% Source PDF page 10
\subsection{Equal Probability Leads to High Entropy}

Consider a BPSK output
\[
x =
\begin{cases}
1, & \text{probability } p \\
0, & \text{probability } 1-p
\end{cases}
\]

\[
H(x) = -\sum_{x=0}^{1} P(x)\log_2 P(x)
\]
\[
= -p\log_2 p - (1-p)\log_2(1-p).
\]
\begin{itemize}
    \item $p$ = probability of symbol $1$
    \item $1-p$ = probability of symbol $0$
    \item $H(x)$ = entropy of the binary source
\end{itemize}

For equal probability $p=0.5$ and $1-p=0.5$,
\[
H(x) = -0.5\log_2 0.5 - 0.5\log_2 0.5 = 1 \text{ bit/symbol.}
\]

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[scale=0.9]
        \draw[->] (0,0) -- (4.6,0) node[right] {$P$};
        \draw[->] (0,0) -- (0,3.1) node[above] {$H(P)$};
        \draw[smooth,domain=0.05:0.95,samples=120,thick,color=chaptercolor]
            plot (\x*4,{(-\x*ln(\x)/ln(2) - (1-\x)*ln(1-\x)/ln(2))*2.8});
        \draw[dashed] (2,0) -- (2,2.8);
        \node[below] at (0,0) {$0$};
        \node[below] at (2,0) {$0.5$};
        \node[below] at (4,0) {$1$};
        \node[left] at (0,2.8) {$1$};
    \end{tikzpicture}
    \caption{Entropy of a binary source versus probability}
\end{figure}

From the figure, it is clear that for $p=0$ or $p=1$, entropy $H(x)$ is zero, but for $p=0.5$, entropy is maximum. Equal probability leads to the highest uncertainty, so it also leads to the highest entropy.

\begin{questionanswerbox}[Example]
Calculate the entropy of the QPSK transmitted signal below:

\[
00\;\;01\;\;10\;\;10\;\;11\;\;11\;\;01\;\;01\;\;10\;\;00\;\;10\;\;10\;\;11\;\;01\;\;10
\]

QPSK has four distinct $(00, 01, 10, 11)$ bit combinations. Thus, $2^n = 4 \Rightarrow n = 2$. So, we have to take two bits for each symbol.

\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{8pt}
\begin{longtable}{|C{2.8cm}|C{2.2cm}|}
    \hline
    \textbf{Symbol} & \textbf{Times} \\ \hline
    \endfirsthead
    \hline
    \textbf{Symbol} & \textbf{Times} \\ \hline
    \endhead
    \endfoot
    00 & 2 \\ \hline
    01 & 4 \\ \hline
    10 & 6 \\ \hline
    11 & 3 \\ \hline
\end{longtable}

\[
P_{00} = \frac{2}{15}, \qquad
P_{01} = \frac{4}{15}, \qquad
P_{10} = \frac{6}{15}, \qquad
P_{11} = \frac{3}{15}
\]

\[
H(X) = -\left[
P_{00}\log_2 P_{00}
+ P_{01}\log_2 P_{01}
+ P_{10}\log_2 P_{10}
+ P_{11}\log_2 P_{11}
\right]
\]
\[
= -\left[
\frac{2}{15}\log_2\frac{2}{15}
+ \frac{4}{15}\log_2\frac{4}{15}
+ \frac{6}{15}\log_2\frac{6}{15}
+ \frac{3}{15}\log_2\frac{3}{15}
\right]
\]
\[
= 1.889 \approx 2 \text{ bits/symbol.}
\]

\textbf{Physical significance:} Thus, we need at least $2$ bits to transmit the signal.
\end{questionanswerbox}

\clearpage

% Source PDF page 11
\section{Joint Entropy}

\begin{definitionbox}[Joint Entropy]
The joint entropy $H(X,Y)$ of a pair of random variables $(X,Y)$ with joint distribution $P(x,y)$ is the average uncertainty associated with the pair of outcomes taken together.
\end{definitionbox}

\begin{formulabox}[Joint Entropy Formula]
\[
H(X,Y) = -\sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(x,y)
\]
\[
= -E\log_2 P(x,y)
\]
\begin{itemize}
    \item $H(X,Y)$ = joint entropy of the variables $X$ and $Y$
    \item $P(x,y)$ = joint probability of the pair $(x,y)$
    \item $E[\cdot]$ = expectation
\end{itemize}
\end{formulabox}

\begin{questionanswerbox}[Example]
Two variables are considered for a student:
\begin{itemize}
    \item $X$ = study hours
    \item $Y$ = attendance
\end{itemize}

Their joint probabilities are listed below. Show that the variables are dependent on each other.

\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{8pt}
\begin{longtable}{|C{2.8cm}|C{2.8cm}|C{2.8cm}|}
    \hline
    \textbf{X} & \textbf{Y} & $\mathbf{P(X,Y)}$ \\ \hline
    \endfirsthead
    \hline
    \textbf{X} & \textbf{Y} & $\mathbf{P(X,Y)}$ \\ \hline
    \endhead
    \endfoot
    Low  & Low  & 0.20 \\ \hline
    Low  & High & 0.10 \\ \hline
    High & Low  & 0.15 \\ \hline
    High & High & 0.55 \\ \hline
\end{longtable}

\textbf{Step 1: Joint entropy contribution of each case}
\begin{align*}
H_{LL}(X,Y) &= -0.2\log_2(0.2) = 0.46 \\
H_{LH}(X,Y) &= -0.1\log_2(0.1) = 0.33 \\
H_{HL}(X,Y) &= -0.15\log_2(0.15) = 0.41 \\
H_{HH}(X,Y) &= -0.55\log_2(0.55) = 0.4743
\end{align*}

\[
H(X,Y) = 0.46 + 0.33 + 0.41 + 0.4743 = 1.6743
\]

\textbf{Step 2: Marginal entropy for study hours}
\[
P_L = 0.2 + 0.1 = 0.3, \qquad P_H = 0.15 + 0.55 = 0.7
\]
\[
H(X) = -\left[0.3\log_2(0.3) + 0.7\log_2(0.7)\right] = 0.881
\]

\textbf{Step 3: Marginal entropy for attendance}
\[
P_L = 0.2 + 0.15 = 0.35, \qquad P_H = 0.1 + 0.55 = 0.65
\]
\[
H(Y) = -\left[0.35\log_2(0.35) + 0.65\log_2(0.65)\right] = 0.934
\]
\end{questionanswerbox}

\begin{infobox}[Conclusion]
\[
H(X) + H(Y) = 0.881 + 0.934 = 1.815
\]
Since $H(X) + H(Y) > H(X,Y)$, the variables $X$ and $Y$ are dependent. We can also say that the variables contain mutual information.
\end{infobox}

\clearpage

% Source PDF page 12
\section{Mutual Information}

\begin{definitionbox}[Mutual Information]
Mutual information measures the amount of information that one random variable $X$ provides about another random variable $Y$. It quantifies the reduction in uncertainty about one variable when the other is known.
\end{definitionbox}

\begin{formulabox}[Mutual Information Formula]
\[
I(X;Y) = H(X) + H(Y) - H(X,Y)
\]
\[
\text{or equivalently,} \qquad I(X;Y) = H(X) - H(X \mid Y)
\]
\begin{itemize}
    \item $I(X;Y)$ = mutual information between $X$ and $Y$
    \item $H(X)$, $H(Y)$ = marginal entropies
    \item $H(X,Y)$ = joint entropy
    \item $H(X \mid Y)$ = conditional entropy of $X$ given $Y$
\end{itemize}
\end{formulabox}

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[scale=1.02, >=stealth]
        \draw[thick, color=chaptercolor] (-1.35,0) circle (1.85);
        \draw[thick, color=chaptercolor] (1.35,0) circle (1.85);
        \fill[softlavender!35] (0,0) ellipse (0.5 and 1.45);

        \node at (-1.85,0) {$H(X \mid Y)$};
        \node at (1.85,0) {$H(Y \mid X)$};
        \node at (0,0) {$I(X;Y)$};

        \node[font=\bfseries] at (0,2.55) {$H(X,Y)$};
        \draw[->, thick] (-0.25,2.28) -- (-0.85,1.55);
        \draw[->, thick] (0.25,2.28) -- (0.85,1.55);

        \node[font=\bfseries] at (-2.95,-2.35) {$H(X)$};
        \draw[->, thick] (-2.55,-2.05) -- (-2.1,-1.55);

        \node[font=\bfseries] at (2.95,-2.35) {$H(Y)$};
        \draw[->, thick] (2.55,-2.05) -- (2.1,-1.55);
    \end{tikzpicture}
    \caption{Mutual information as the overlap between $H(X)$ and $H(Y)$}
\end{figure}

\begin{questionanswerbox}[Example]
Calculate the joint entropy for the following data and also find the mutual information.

\renewcommand{\arraystretch}{1.8}
\setlength{\tabcolsep}{8pt}
\begin{longtable}{|C{3.4cm}|C{2.4cm}|C{2.4cm}|}
    \hline
    \textbf{Y $\backslash$ X} & 0 & 1 \\ \hline
    \endfirsthead
    \hline
    \textbf{Y $\backslash$ X} & 0 & 1 \\ \hline
    \endhead
    \endfoot
    $0$ & $\frac{1}{2}$ & $\frac{1}{4}$ \\ \hline
    $1$ & $\frac{1}{8}$ & $\frac{1}{8}$ \\ \hline
\end{longtable}

\textbf{Step 1: Joint entropy}
\begin{align*}
H(X,Y) &= -\left[\frac{1}{2}\log_2\frac{1}{2} + \frac{1}{4}\log_2\frac{1}{4}
+ \frac{1}{8}\log_2\frac{1}{8} + \frac{1}{8}\log_2\frac{1}{8}\right] \\
&= 1.75
\end{align*}

\textbf{Step 2: Entropy of $Y$}
\[
P(Y=0) = \frac{1}{2} + \frac{1}{4} = \frac{3}{4}, \qquad
P(Y=1) = \frac{1}{8} + \frac{1}{8} = \frac{1}{4}
\]
\[
H(Y) = \frac{3}{4}\log_2\frac{4}{3} + \frac{1}{4}\log_2 4 = 0.811
\]

\textbf{Step 3: Entropy of $X$}
\[
P(X=0) = \frac{1}{2} + \frac{1}{8} = \frac{5}{8}, \qquad
P(X=1) = \frac{1}{4} + \frac{1}{8} = \frac{3}{8}
\]
\[
H(X) = \frac{5}{8}\log_2\frac{8}{5} + \frac{3}{8}\log_2\frac{8}{3} = 0.954
\]

\textbf{Step 4: Conditional entropies}
\[
H(X \mid Y) = H(X,Y) - H(Y) = 1.75 - 0.811 = 0.939
\]
\[
H(Y \mid X) = H(X,Y) - H(X) = 1.75 - 0.954 = 0.796
\]

\textbf{Step 5: Mutual information}
\[
I(X;Y) = H(X) + H(Y) - H(X,Y) = 0.954 + 0.811 - 1.75 = 0.015
\]
\end{questionanswerbox}

\clearpage

% Source PDF page 13
\section{Conditional Entropy}

\begin{definitionbox}[Conditional Entropy]
Conditional entropy measures the amount of uncertainty remaining in a random variable when another related variable is already known. It tells us how much additional information is still needed to describe X when Y is already provided.
\end{definitionbox}

\subsection{Simple Intuition}
Consider the set $\{1,2,3,4,5,6\}$. If we only know whether the outcome is odd or even, some uncertainty still remains.
\begin{itemize}
    \item If the outcome is \textbf{odd}, the possible values are $\{1,3,5\}$.
    \item If the outcome is \textbf{even}, the possible values are $\{2,4,6\}$.
\end{itemize}
This remaining uncertainty is the idea captured by conditional entropy.

\begin{formulabox}[Conditional Entropy Formula]
\[
H(Y \mid X) = -\sum_{x \in X} P(x)\sum_{y \in Y} P(y \mid x)\log_2 P(y \mid x)
\]
\[
= -\sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(y \mid x)
\]
\begin{itemize}
    \item $H(Y \mid X)$ = uncertainty of $Y$ after $X$ is known
    \item $P(y \mid x)$ = conditional probability of $Y=y$ given $X=x$
    \item $P(x,y)$ = joint probability of $X$ and $Y$
\end{itemize}
\end{formulabox}

\begin{formulabox}[Useful Relation for the Derivation]
\[
P(y \mid x)=\frac{P(x,y)}{P(x)}
\]
\[
\log(AB)=\log A+\log B
\]
\begin{itemize}
    \item The first relation rewrites conditional probability in terms of joint and marginal probability.
    \item The logarithm identity is used when expanding \(\log_2\bigl(P(x)P(y \mid x)\bigr)\).
\end{itemize}
\end{formulabox}

\subsection{Relation with Joint Entropy}
Starting from the joint entropy,
\begin{align*}
H(X,Y) &= -\sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(x,y) \\
&= -\sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2\bigl(P(x)P(y \mid x)\bigr) \\
&= -\sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(x)
- \sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(y \mid x) \\
&= -\sum_{x \in X} P(x)\log_2 P(x)
- \sum_{x \in X}\sum_{y \in Y} P(x,y)\log_2 P(y \mid x)
\end{align*}

Hence,
\[
H(X,Y) = H(X) + H(Y \mid X)
\]

Similarly,
\[
H(X,Y) = H(Y) + H(X \mid Y)
\]

\clearpage

% Source PDF page 14
\section{Relative Entropy}

\begin{definitionbox}[Relative Entropy]
Relative entropy, also called Kullback-Leibler divergence, measures how different an estimated probability distribution is from the true distribution.
\end{definitionbox}

\begin{infobox}[What Relative Entropy Means]
\begin{itemize}
    \item Relative entropy compares two probability mass functions: \(P(x)\) and \(Q(x)\).
    \item \(P(x)\) represents the true or actual distribution.
    \item \(Q(x)\) represents the estimated or assumed distribution.
    \item It tells us how far the estimate \(Q(x)\) is from the true distribution \(P(x)\).
    \item A larger value of \(D(P \parallel Q)\) means the deviation from the actual distribution is higher.
    \item If the two distributions are identical, then the relative entropy becomes zero.
\end{itemize}
\end{infobox}

\begin{formulabox}[Kullback-Leibler Divergence]
\[
D(P \parallel Q) = \sum_{x \in X} P(x)\log_2 \frac{P(x)}{Q(x)}
\]
\begin{itemize}
    \item $P(x)$ = true probability mass function
    \item $Q(x)$ = estimated probability mass function
    \item $D(P \parallel Q)$ = relative entropy from $Q$ to $P$
\end{itemize}
\end{formulabox}

\begin{infobox}[Basic Points]
\begin{itemize}
    \item If $P(x) = Q(x)$, then $D(P \parallel Q) = 0$.
        \begin{figure}[H]
        \centering
        \begin{tikzpicture}[x=1.7cm,y=2.1cm]
            \draw[->, thick] (0,0) -- (0,1.45);
            \draw[->, thick] (0,0) -- (3.4,0);

            \draw[thick] (0.45,0) rectangle (0.95,1.1);
            \draw[thick] (1.55,0) rectangle (2.05,0.8);
            \draw[thick] (2.65,0) rectangle (3.15,0.68);

            \node[above] at (0.7,1.1) {$P$};
            \node[above] at (1.8,0.8) {$A$};
            \node[above] at (2.9,0.68) {$B$};
        \end{tikzpicture}
        \caption{Relative entropy illustration comparing the true distribution $P$ with estimates $A$ and $B$}
        \end{figure}
    \item In the illustration:
    \[
    D(P \parallel P) = 0, \qquad D(P \parallel A) = 0.25, \qquad D(P \parallel B) = 1.25
    \]
    \item Higher relative entropy means the estimated distribution deviates more from the actual distribution.
\end{itemize}
\end{infobox}


\begin{questionanswerbox}[Example]
Consider a scenario where the true distribution is
\[
P_{\text{dog}} = 1, \qquad P_{\text{cat}} = 0
\]

For model $A$,
\[
P'_{\text{dog}} = 0.1, \qquad P'_{\text{cat}} = 0.9
\]

For model $B$,
\[
P'_{\text{dog}} = 0.9, \qquad P'_{\text{cat}} = 0.1
\]

Find the relative entropy using KL divergence.

\textbf{For model $A$:}
\[
D(P \parallel P'_A)
= P_{\text{dog}}\log_2\frac{P_{\text{dog}}}{P'_{\text{dog}}}
+ P_{\text{cat}}\log_2\frac{P_{\text{cat}}}{P'_{\text{cat}}}
\]
\[
= 1\log_2\frac{1}{0.1} + 0\cdot\log_2\frac{0}{0.9}
= \log_2 10
= 3.32 \text{ unit}
\]

\textbf{For model $B$:}
\[
D(P \parallel P'_B)
= P_{\text{dog}}\log_2\frac{P_{\text{dog}}}{P'_{\text{dog}}}
+ P_{\text{cat}}\log_2\frac{P_{\text{cat}}}{P'_{\text{cat}}}
\]
\[
= 1\log_2\frac{1}{0.9} + 0\cdot\log_2\frac{0}{0.1}
= 0.15 \text{ unit}
\]

So model $B$ is better because it has lower KL divergence.
\end{questionanswerbox}

\clearpage

% Source PDF page 15
\subsection{Mutual Information as Relative Entropy}
For two dependent random variables $(X,Y)$, mutual information can be written as the relative entropy between the joint distribution $P(x,y)$ and the product of the marginals $P(x)P(y)$:

\begin{formulabox}[Mutual Information from Relative Entropy]
\[
I(X;Y) = \sum_{x,y} P(x,y)\log_2 \frac{P(x,y)}{P(x)P(y)}
= D\bigl(P(x,y)\,\|\,P(x)P(y)\bigr)
\]
\begin{itemize}
    \item $I(X;Y)$ = mutual information between $X$ and $Y$
    \item $P(x,y)$ = joint probability distribution of $X$ and $Y$
    \item $P(x)$, $P(y)$ = marginal probability distributions of $X$ and $Y$
    \item $P(x)P(y)$ = product of marginals, representing the independent case
    \item $D\bigl(P(x,y)\,\|\,P(x)P(y)\bigr)$ = relative entropy between the joint distribution and the product of marginals
\end{itemize}
\end{formulabox}

\begin{align*}
I(X;Y)
&= \sum_{x,y} P(x,y)\log_2 \frac{P(x,y)}{P(x)P(y)}\\
&= \sum_{x,y} P(x,y)\log_2 \frac{P(x)P(y \mid x)}{P(x)P(y)} \\
&= \sum_{x,y} P(x,y)\log_2 \frac{P(y \mid x)}{P(y)} \\
&= \sum_{x,y} P(x,y)\log_2 P(y \mid x) - \sum_{x,y} P(x,y)\log_2 P(y) \\
&= -H(Y \mid X) + H(Y)\\
\text{So, } I(X;Y) &= H(Y) - H(Y \mid X)
\end{align*}

Also,
\[
H(X,Y) = H(X) + H(Y \mid X)
\]

Therefore,
\[
H(Y \mid X) = H(X,Y) - H(X)
\]

Substituting this into the previous expression,
\[
I(X;Y) = H(Y) - H(X,Y) + H(X)
\]

Thus the mutual information formula becomes
\[
I(X;Y) = H(X) + H(Y) - H(X,Y)
\]

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[scale=1.02, >=stealth]
        \draw[thick, color=chaptercolor] (-1.35,0) circle (1.85);
        \draw[thick, color=chaptercolor] (1.35,0) circle (1.85);
        \fill[softlavender!35] (0,0) ellipse (0.5 and 1.45);

        \node at (-1.85,0) {$H(X \mid Y)$};
        \node at (1.85,0) {$H(Y \mid X)$};
        \node at (0,0) {$I(X;Y)$};

        \node[font=\bfseries] at (0,2.55) {$H(X,Y)$};
        \draw[->, thick] (-0.25,2.28) -- (-0.85,1.55);
        \draw[->, thick] (0.25,2.28) -- (0.85,1.55);

        \node[font=\bfseries] at (-2.95,-2.35) {$H(X)$};
        \draw[->, thick] (-2.55,-2.05) -- (-2.1,-1.55);

        \node[font=\bfseries] at (2.95,-2.35) {$H(Y)$};
        \draw[->, thick] (2.55,-2.05) -- (2.1,-1.55);
    \end{tikzpicture}
    \caption{Mutual information as the overlap between $H(X)$ and $H(Y)$}
\end{figure}

\clearpage

% Source PDF page 16
\subsection{Convexity and Jensen's Inequality}

\begin{definitionbox}[Convex Function]
A function $f$ is convex over an interval if its second derivative is non-negative over that interval:
\[
\frac{d^2 f(x)}{dx^2} \ge 0
\]
Examples: $x^2$ and $e^x$.
\end{definitionbox}

Consider the function $f(x)=x^2$ with $a=0$ and $b=2$.

\begin{itemize}
    \item $f(0)=0$ and $f(2)=4$
    \item Average of the function values:
\[
\frac{f(0)+f(2)}{2} = \frac{0+4}{2} = 2
\]
    \item Function of the average:
\[
f\left(\frac{a+b}{2}\right)
= f\left(\frac{0+2}{2}\right)
= f(1)
= 1^2
= 1
\]
\end{itemize}

\begin{infobox}[Key Observation]
For a convex function, the average of the function is greater than or equal to the function of the average:
\[
\text{Average of the function} \ge \text{Function of the average}
\]
\end{infobox}

\begin{formulabox}[Jensen's Inequality]
If a function $f(x)$ is convex over an interval $(a,b)$, then for every $x_1,x_2 \in (a,b)$ and $0 \le \lambda \le 1$,
\[
\lambda f(x_1) + (1-\lambda)f(x_2) \ge f\bigl[\lambda x_1 + (1-\lambda)x_2\bigr]
\]
A function $f$ is said to be strictly convex if equality holds only when $\lambda = 0$ or $\lambda = 1$.
\end{formulabox}

\begin{questionanswerbox}[Example]
For $x_1=0$, $x_2=2$, $\lambda=0.5$, and $f(x)=x^2$,
\[
\lambda f(x_1) + (1-\lambda)f(x_2) \ge f\bigl[\lambda x_1 + (1-\lambda)x_2\bigr]
\]
\[
0.5\times 0 + 0.5\times 4 \ge f[0.5\times 0 + 0.5\times 2]
\]
\[
2 \ge 1
\]
Thus $f(x)=x^2$ is a convex function.
\end{questionanswerbox}

\clearpage

% Source PDF page 17
\subsection{Gibbs Information Inequality}

\begin{infobox}[Statement]
Let $P(x)$ and $Q(x)$, where $x \in X$, be two probability mass functions. Then
\[
D(P \parallel Q) \ge 0
\]
with equality if and only if $P(x)=Q(x)$ for all $x$.
\end{infobox}

Let
\[
A = \{x : P(x) > 0\}
\]
be the support set of $P(x)$. Then
\begin{align*}
-D(P \parallel Q)
&= -\sum_{x\in A} P(x)\log \frac{P(x)}{Q(x)} \\
&= \sum_{x\in A} P(x)\log \frac{Q(x)}{P(x)} \\
&\le \log \sum_{x\in A} P(x)\frac{Q(x)}{P(x)} \\
&= \log \sum_{x\in A} Q(x) \\
&\le \log \sum_{x\in X} Q(x) \\
&= \log 1 \\
&= 0
\end{align*}

The inequality step follows because $\log(\cdot)$ is a strictly concave function.

Therefore,
\[
D(P \parallel Q) \ge 0
\]
and relative entropy becomes zero only when the two distributions are identical.

\clearpage

% Source PDF page 18
\subsection{Example: Asymmetric Nature of Relative Entropy}

\begin{questionanswerbox}[Example]
Let $X=\{0,1\}$ and consider two distributions $P$ and $Q$ on $X$:
\[
P(0)=1-r,\qquad P(1)=r,\qquad Q(0)=1-s,\qquad Q(1)=s
\]
Find $D(P \parallel Q)$ and $D(Q \parallel P)$.

We have
\[
D(P \parallel Q)
= \sum_x P(x)\log_2 \frac{P(x)}{Q(x)}
= P(0)\log_2\frac{P(0)}{Q(0)} + P(1)\log_2\frac{P(1)}{Q(1)}
\]
\[
= (1-r)\log_2\frac{1-r}{1-s} + r\log_2\frac{r}{s}
\]

When $r=s=\dfrac{1}{2}$, then
\[
D(P \parallel Q)=0
\]

When $r=\dfrac{1}{2}$ and $s=\dfrac{1}{4}$,
\[
D(P \parallel Q)
= \frac{1}{2}\log_2\frac{1-\frac{1}{2}}{1-\frac{1}{4}}
+ \frac{1}{2}\log_2\frac{\frac{1}{2}}{\frac{1}{4}}
\]
\[
= \frac{1}{2}\log_2\frac{\frac{1}{2}}{\frac{3}{4}}
+ \frac{1}{2}\log_2\frac{\frac{1}{2}}{\frac{1}{4}}
\]
\[
= -0.292481 + \frac{1}{2}
= 0.2075 \text{ bit}
\]

Similarly,
\[
D(Q \parallel P)
= (1-s)\log_2\frac{1-s}{1-r} + s\log_2\frac{s}{r}
\]
\[
= \left(1-\frac{1}{4}\right)\log_2\frac{1-\frac{1}{4}}{1-\frac{1}{2}}
+ \frac{1}{4}\log_2\frac{\frac{1}{4}}{\frac{1}{2}}
\]
\[
= \frac{3}{4}\log_2\frac{\frac{3}{4}}{\frac{1}{2}}
+ \frac{1}{4}\log_2\frac{1}{2}
\]
\[
= 0.438721 - 0.25
\]
\[
= 0.188721 \text{ bit}
\]

Hence,
\[
D(P \parallel Q) \ne D(Q \parallel P)
\]
So the asymmetric relationship is proved.
\end{questionanswerbox}

\clearpage

% Source PDF page 19
\section{Chain Rule}

\subsection{Entropy}
\begin{formulabox}[Chain Rule for Entropy]
\[
H(X_1,X_2,\ldots,X_n) = \sum_{i=1}^{n} H(X_i \mid X_1,\ldots,X_{i-1})
\]
\end{formulabox}

\begin{itemize}
    \item For $n=2$:
\[
H(1,2)=H(1)+H(2 \mid 1)
\]
    \item For $n=3$:
\[
H(1,2,3)=H(1)+H(2 \mid 1)+H(3 \mid 1,2)
\]
\end{itemize}

\subsection{Conditional Entropy}
\begin{formulabox}[Chain Rule for Conditional Entropy]
\[
H(X_1,X_2,\ldots,X_n \mid Y) = \sum_{i=1}^{n} H(X_i \mid X_1,\ldots,X_{i-1},Y)
\]
\end{formulabox}

\subsection{Mutual Information}
\begin{formulabox}[Chain Rule for Mutual Information]
\[
I(X_1,X_2,\ldots,X_n;Y) = \sum_{i=1}^{n} I(X_i;Y \mid X_1,\ldots,X_{i-1})
\]
\end{formulabox}

\begin{itemize}
    \item For $n=2$:
\[
I(1,2;Y)=I(1;Y)+I(2;Y \mid 1)
\]
    \item For $n=3$:
\[
I(1,2,3;Y)=I(1;Y)+I(2;Y \mid 1)+I(3;Y \mid 1,2)
\]
\end{itemize}

\subsection{Worked Examples}

\begin{questionanswerbox}[Entropy of a Finite-Length DMS]
Calculate the entropy of the finite-length DMS $X$ with probability distribution
\[
P(x)=\{0.125,0.25,0.5,0.125\}, \qquad x \in X
\]

Entropy,
\[
H(X) = \sum_{x\in X} H(x_i) = -\sum_{i=1}^{n} p_i \log_2 p_i
\]
\[
= -\left[2(0.125)\log_2(0.125) + (0.25)\log_2(0.25) + (0.5)\log_2(0.5)\right]
\]
\[
= 1.375
\]
\end{questionanswerbox}

\begin{questionanswerbox}[Average Self Information of a Source]
For any source $X$ with pmf
\[
P(x)=\{0.0625,0.5,0.25,0.125,0.0625\}, \qquad x \in X
\]
calculate its average self information.

The average self information is nothing but the entropy.

\[
H(X) = -\left[2\times 0.0625\log_2 0.0625 + 0.5\log_2 0.5 + 0.25\log_2 0.25 + 0.125\log_2 0.125\right]
\]
\[
= 1.875
\]
\end{questionanswerbox}

\clearpage

% Source PDF page 20
\section{Worked Problems}

\begin{questionanswerbox}[Fully Dependent System]
Prove that for a fully dependent system, mutual information is the same as self information.

Let two random variables $X$ and $Y$ be fully dependent. Knowing $X$ gives full knowledge of $Y$, and vice versa. Therefore,
\[
H(Y \mid X)=0
\]

From the formula of mutual information,
\[
I(X;Y)=H(Y)-H(Y \mid X)
\]
\[
= H(Y)-0
\]

So,
\[
I(X;Y)=H(Y)
\]

Hence, mutual information equals the entropy, that is, the average self information. All uncertainty in one variable is resolved by knowing the other.
\end{questionanswerbox}

\begin{questionanswerbox}[8-PSK Modulator Example]
An 8-PSK modulator generates the following symbol sequence:
\[
010\;\;110\;\;111\;\;101\;\;011\;\;000\;\;011\;\;101\;\;110
\]
\[
101\;\;110\;\;111\;\;010\;\;001\;\;101\;\;100
\]
Determine the average information of the modulator.

The average information is the entropy.

\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{8pt}
\begin{longtable}{|C{2.8cm}|C{2.3cm}|}
    \hline
    \textbf{Symbol} & \textbf{Count} \\ \hline
    \endfirsthead
    \hline
    \textbf{Symbol} & \textbf{Count} \\ \hline
    \endhead
    \endfoot
    000 & 1 \\ \hline
    001 & 1 \\ \hline
    010 & 2 \\ \hline
    011 & 2 \\ \hline
    100 & 1 \\ \hline
    101 & 4 \\ \hline
    110 & 3 \\ \hline
    111 & 2 \\ \hline
\end{longtable}

\[
H = -\left[
3\times \frac{1}{16}\log_2\frac{1}{16}
+ 3\times \frac{2}{16}\log_2\frac{2}{16}
+ \frac{3}{16}\log_2\frac{3}{16}
+ \frac{4}{16}\log_2\frac{4}{16}
\right]
\]
\[
= \frac{3}{4} + \frac{9}{8} + 0.4528195311 + \frac{1}{2}
\]
\[
\approx 2.8278 \text{ bits/symbol}
\]

\textbf{Significance:} At least $3$ bits are required to transmit the 8-PSK modulated signal.
\end{questionanswerbox}

% Source PDF page 21
\section{Channel Interpretation}

\begin{questionanswerbox}[Mutual Information as a Measure of Channel Goodness]
Mutual information can be interpreted as a measure of channel goodness. Justify the statement with an example.

\paragraph*{Answer}
\begin{itemize}
    \item Mutual information \(I(X;Y)\) quantifies how much information the received signal \(Y\) reveals about the transmitted signal \(X\).
    \item A higher mutual information means less uncertainty in decoding and a more reliable channel.
    \item For a perfect channel, \(I(X;Y)=H(X)\). All transmitted information is received without loss.
    \item For a noisy channel, \(I(X;Y)=0\). The output reveals nothing about the input, indicating a useless or fully noisy channel.
\end{itemize}

\paragraph*{Example}
For a binary symmetric channel with low crossover probability \(p\), \(I(X;Y)\) is high, showing good channel reliability. For BSC, let \(p=0.1\).

\[
C=\max_{P(X)} I(X;Y)=1-H(p)
\]
\[
= 1-\left[-p\log_2 p-(1-p)\log_2(1-p)\right]
\]
\[
= 1+0.1\log_2(0.1)+0.9\log_2(0.9)
\]
\[
= 0.531 \text{ bit per bit}
\]

The channel is fairly good but not perfect.
\end{questionanswerbox}

\begin{questionanswerbox}[Probability of Correct Reception]
Prove that the probability of correct reception is \(1-P_e\).

\paragraph*{Answer}
Let \(P_c\) be the probability of correct reception and \(P_e\) be the probability of error.

Since these are complementary events,
\[
P_c+P_e=1
\]
or,
\[
P_c=1-P_e
\]

Hence proved.
\end{questionanswerbox}

% Source PDF page 22
\section{Discrete Memoryless Channel}

\begin{definitionbox}[Discrete Memoryless Channel (DMC)]
A discrete memoryless channel (DMC) is a channel where:
\begin{itemize}
    \item the input and output alphabets are discrete (finite),
    \item the current output depends only on the current input,
    \item the current output does not depend on any previous inputs or outputs.
\end{itemize}
So, the channel has no memory.
\end{definitionbox}

\begin{definitionbox}[Channel Matrix]
The channel matrix of a DMC is a table showing the conditional probabilities of receiving output symbols given an input symbol.
\end{definitionbox}

\begin{questionanswerbox}[Example]
Let the input alphabet be \(\{0,1\}\) and the output alphabet be \(\{0,1\}\). If
\[
P(y=0\mid x=0)=0.9,\qquad P(y=1\mid x=0)=0.1
\]
\[
P(y=1\mid x=1)=0.8,\qquad P(y=0\mid x=1)=0.2
\]
then the channel matrix is
\[
\begin{bmatrix}
0.9 & 0.1 \\
0.2 & 0.8
\end{bmatrix}
\]
\end{questionanswerbox}

\subsection{Average Information Rate of a Source}

\begin{questionanswerbox}[Worked Example]
The output of an information source consists of \(150\) symbols, \(32\) of which occur with a probability of \(\frac{1}{64}\), and the remaining with a probability of \(\frac{1}{236}\). The source emits \(2000\) symbols/sec. Assume the symbols are chosen independently. Find the average information rate of this source.

\paragraph*{Given}
\begin{itemize}
    \item \(p_1=\frac{1}{64}\)
    \item \(p_2=\frac{1}{236}\)
    \item \(n_1=32\)
    \item \(n_2=150-32=118\)
\end{itemize}

\paragraph*{Total Entropy}
\[
H(X)=-\left(p_1\log_2 p_1 \times n_1 + p_2\log_2 p_2 \times n_2\right)
\]
\[
= -\left(\frac{1}{64}\log_2\frac{1}{64}\times 32 + \frac{1}{236}\log_2\frac{1}{236}\times 118\right)
\]
\[
= 6.9413 \text{ bit/symbol}
\]

\paragraph*{Average Information Rate}
\[
R=H(X)\times r
\]
\[
= 6.9413\times 2000
\]
\[
= 13882.64305 \text{ bits/sec}
\]
\end{questionanswerbox}

% Source PDF page 23
\section{Additional Worked Problems}

\begin{questionanswerbox}[Maximum Entropy for Two Messages]
Consider a source that transmits two independent messages with probabilities \(p\) and \((1-p)\), respectively. Demonstrate that the entropy is maximized when the two messages are equally likely. Plot the entropy of source vs probability.

\paragraph*{Answer}
For equally likely messages, the probabilities are
\[
p = 1-p = 0.5
\]

Entropy,
\[
H(p)=-p\log_2 p-(1-p)\log_2(1-p)
\]
\[
= -0.5\log_2 0.5 - 0.5\log_2 0.5
\]
\[
= 1
\]

Hence, the entropy is maximum when \(p=0.5\). When both terms are equally probable, the uncertainty is highest; hence the entropy is maximum. This represents the most information per message.
\end{questionanswerbox}

\begin{questionanswerbox}[B\&W TV Picture Information Rate]
A B\&W TV picture consists of \(525\) lines of picture information. Assume that each element can have \(256\) brightness levels. If pictures are repeated at a rate of \(32\) frame/sec, calculate the average rate of information conveyed by a TV set to a viewer.

\paragraph*{Given}
\begin{itemize}
    \item Number of lines per frame \(=525\)
    \item Brightness levels per picture element \(=256\)
    \item Frame rate \(=32\) fps
    \item Pixels per line \(=256\) (as used in the note)
\end{itemize}

\paragraph*{Pixels per Frame}
\[
525\times 256 = 13440 \text{ pixels/frame}
\]

\paragraph*{Bits per Pixel}
\[
\log_2 256 = 8 \text{ bits}
\]

\paragraph*{Bits per Frame}
\[
8\times 13440 = 107520 \text{ bits/frame}
\]

\paragraph*{Information Rate}
\[
107520\times 32 = 3440640 \text{ bps}
\]
\[
= 3.44 \text{ Mbps}
\]
\end{questionanswerbox}

% Source PDF pages 24-25 are unrelated inserts in the source PDF and are intentionally skipped.

% Source PDF page 26
\begin{questionanswerbox}[Finding \(H(X)\), \(H(Y)\), \(H(X \mid Y)\), and \(H(Y \mid X)\)]
Calculate \(H(X)\), \(H(Y)\), \(H(X \mid Y)\), and \(H(Y \mid X)\) from the data given below.

\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{6pt}
\begin{longtable}{|C{2cm}|C{1.6cm}|C{1.6cm}|C{1.6cm}|C{1.6cm}|}
    \hline
    \textbf{Y $\backslash$ X} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} \\ \hline
    \endfirsthead
    \hline
    \textbf{Y $\backslash$ X} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} \\ \hline
    \endhead
    \endfoot
    1 & \(\dfrac{1}{8}\)  & \(\dfrac{1}{16}\) & \(\dfrac{1}{32}\) & \(\dfrac{1}{32}\) \\ \hline
    2 & \(\dfrac{1}{16}\) & \(\dfrac{1}{8}\)  & \(\dfrac{1}{32}\) & \(\dfrac{1}{32}\) \\ \hline
    3 & \(\dfrac{1}{16}\) & \(\dfrac{1}{16}\) & \(\dfrac{1}{16}\) & \(\dfrac{1}{16}\) \\ \hline
    4 & \(\dfrac{1}{4}\)  & 0                 & 0                 & 0                 \\ \hline
\end{longtable}

From the table above,
\[
P(x_1)=\frac{1}{8}+\frac{1}{16}+\frac{1}{16}+\frac{1}{4}=\frac{1}{2}
\]
\[
P(x_2)=\frac{1}{16}+\frac{1}{8}+\frac{1}{16}+0=\frac{1}{4}
\]
\[
P(x_3)=\frac{1}{32}+\frac{1}{32}+\frac{1}{16}=\frac{1}{8}
\]
\[
P(x_4)=\frac{1}{32}+\frac{1}{32}+\frac{1}{16}=\frac{1}{8}
\]

We have,
\[
H(X)=-\sum_{i=1}^{4} P(x_i)\log_2 P(x_i)
\]
\[
= \frac{1}{2}\log_2 2 + \frac{1}{4}\log_2 2^2 + \frac{1}{8}\log_2 2^3 + \frac{1}{8}\log_2 2^3
\]
\[
= \frac{1}{2}+\frac{1}{2}+\frac{3}{8}+\frac{3}{8}
\]
\[
= \frac{7}{4} \text{ bits/symbol}
\]

Now,
\[
P(y_1)=\frac{1}{8}+\frac{1}{16}+\frac{1}{32}+\frac{1}{32}=\frac{1}{4}
\]
\[
P(y_2)=\frac{1}{16}+\frac{1}{8}+\frac{1}{32}+\frac{1}{32}=\frac{1}{4}
\]
\[
P(y_3)=\frac{4}{16}=\frac{1}{4}
\]
\[
P(y_4)=\frac{1}{4}
\]

% Source PDF page 27
Now,
\[
H(Y)=-\sum_{i=1}^{4} P(y_i)\log_2 P(y_i)
\]
\[
= \frac{4}{4}\log_2 4
\]
\[
= \log_2 2^2
\]
\[
= 2 \text{ bits/symbol}
\]

Now, joint entropy,
\[
H(X,Y)=-\sum_{j=1}^{4}\sum_{i=1}^{4} P(x_i,y_j)\log_2 P(x_i,y_j)
\]
\[
= \frac{2}{8}\log_2 8 + \frac{6}{16}\log_2 16 + \frac{4}{32}\log_2 32 + \frac{1}{4}\log_2 4
\]
\[
= \frac{2}{8}\log_2 2^3 + \frac{6}{16}\log_2 2^4 + \frac{4}{32}\log_2 2^5 + \frac{1}{4}\log_2 2^2
\]
\[
= \frac{6}{8}\log_2 2 + \frac{24}{16}\log_2 2 + \frac{20}{32}\log_2 2 + \frac{2}{4}\log_2 2
\]
\[
= \frac{6}{8} + \frac{24}{16} + \frac{20}{32} + \frac{1}{2}
\]
\[
= \frac{27}{8} \text{ bits/symbol}
\]

Now,
\[
H(X \mid Y)=H(X,Y)-H(Y)=\frac{27}{8}-2=1.375 \text{ bits/symbol}
\]

and,
\[
H(Y \mid X)=H(X,Y)-H(X)=\frac{27}{8}-\frac{7}{4}=1.625 \text{ bits/symbol}
\]
\end{questionanswerbox}

\end{document}
